Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-13 report/report.tex
updated

\begin{algorithm}
\caption{Regionally Additive Models (RAMs)}
\label{alg:ram}
\begin{algorithmic}[1]
\Require Dataset \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\), black-box model \(f(\mathbf{x})\), regional effect method \(\mathcal{R}\), RAM \(f^{\mathtt{RAM}}(\mathbf{x})\)
\State Fit a black-box model \(f(\mathbf{x})\) to \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\)
\State Identify the subregions \(\mathcal{R}_{sk}\) using a regional effect method
\State Define the RAM \(f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xcc \in \mathcal{R}_{sk}}\)
\State Fit the univariate functions \(f_{sk}(x_s)\) for each \(s\) and \(k\).
\end{algorithmic}
\end{algorithm}

RAM is a framework for fitting explainable-by-design models that are expressive enough to capture any interaction.
This is achieved using a pipeline that consists of a sequence of steps (Algorithm~\ref{alg:ram}).
In the first step, we fit a black-box model \(f\) to the data.
The black-box model can be any model that is expressive enough to accurately fit the data.
For example, we can use a random forest or a neural network.
In the second step, we identify the subregions where the black-box model is nearly locally additive.
This is achieved using a regional effect method \(\mathcal{R}\).
For example, we can use REPID (add citation) or any method presented by % \citet{ribeiro2016should}.
At section \ref{sec:regional_effect_methods} we present a regional effect method based on the ALE plot.
This step defines the form of the final model as defined in equation \ref{eq:ram}.
Finally, we fit a GAM to the identified subregions, using any GAM model.

@@

\begin{equation}  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \underbrace{\frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
    \mathcal{S}_k} \dfdx(\mathbf{x}^i)}_{\hat{\mu}(z_{k-1}, z_k)})
\end{equation}
%
where \(k_x\) the index of the bin such that
\(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
is the set of the instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\).
The average effect is computed by dividing the feature space into bins and computing the average effect of
the feature \(x_s\) in each bin.
In cases where there are interactions between the features, the instance-level effects inside each bin
start to deviate from the average effect.
We can measure such deviation using the standard deviation of the instance-level effects inside each bin:
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}
%
and afterwards, the level of interaction between the feature \(x_s\) and the rest of the features can be quantified
as the aggregated bin-deviation \(\sqrt{\sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k)}\).
Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
that minimize the aggregated bin-deviation defined on the subregions.
For this reason we define the bin-effect and bin-deviation on a subregion \(\mathcal{R}_{st}\) as:

@@

\begin{equation}
  \label{eq:bin_deviation_subregion}
  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
    \hat{\mu}(\mathcal{R}_{st}) \right )^2
\end{equation}

Therefore, the interaction of the feature \(x_s\) with the rest of the features within a subregion \(\mathcal{R}_{st}\)
can be quantified as:

\begin{equation}
  \label{eq:SE}
  \mathcal{L}(\mathcal{R}_{sk}) = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
\end{equation}

@@

\begin{algorithm}
\caption{DALE-based Region Detection}
\label{alg:subregion_identification}
\begin{algorithmic}
\Procedure{DALE\_Region\_Detection}{$\mathcal{D}$, $J$, $T$}
    \For {\(s=1, \cdots, D\)}
        \State $\mathtt{D\_before\_split = [D]}$
        \For {\(t=1, \cdots, T\)}
        \EndFor
    \EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{DALE-based Region Detection}
\label{alg:single_feature_subregion}
\begin{algorithmic}
\For{\( c = 1, \cdots, s-1, s+1, \cdots, D\)}
    \State Lalala
    \State $\mathtt{D\_before\_split = [D]}$
    \State $\mathtt{L\_before\_split = [}\mathcal{L}(\mathcal{D}_{s1}) \mathtt{]}$
    \State $\mathtt{L\_after\_split = []}$
\EndFor
\end{algorithmic}
\end{algorithm}

@@

>>>>

@@

\begin{equation}  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \underbrace{\frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
    \mathcal{S}_k} \dfdx(\mathbf{x}^i)}_{\hat{\mu}(z_{k-1}, z_k)})
\end{equation}
%
where \(k_x\) the index of the bin such that
\(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
is the set of the instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\).
The average effect is computed by dividing the feature space into bins and computing the average effect of
the feature \(x_s\) in each bin.
In cases where there are interactions between the features, the instance-level effects inside each bin
start to deviate from the average effect.
We can measure such deviation using the standard deviation of the instance-level effects inside each bin:
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}
%
and the global interaction between the feature \(x_s\) and the rest of the features with the
aggregated bin-deviation \({\left ( \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) \right)}^{\frac{1}{2}} \).
Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
that minimize the aggregated bin-deviation defined on the subregions.
For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:

@@

\begin{equation}
  \label{eq:bin_deviation_subregion}
  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
    \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}

\begin{equation}
  \label{eq:SE}
  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
\end{equation}

The goal is to find the optimal set of subregions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) that minimize
the aggregated deviation. For this reason, we define the following optimization problem:

\begin{equation}
  \label{eq:optimal_subregions}
  \begin{aligned}
    & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
    & & \sum_{t=1}^{T_s} \mathcal{L}_{\mathcal{R}_{st}} \\
    & \text{subject to}
    & & \bigcup_{t=1}^{T_s} \mathcal{R}_{st} = \mathcal{X} \\
    & & & \mathcal{R}_{st} \cap \mathcal{R}_{s\tau} = \emptyset, \quad \forall t \neq \tau
  \end{aligned}
\end{equation}

@@

\begin{algorithm}
\caption{DALE-based Region Detection}
\label{alg:subregion_identification}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{X, J, T}
\BlankLine
\Output{optimal\_splits}
\BlankLine
splits[s, c, t] $\gets$ None\;
positions[s, c, t] $\gets$ None\;
\For{$s = 1$ to $D$}{
    \For{$c \in \{1, \dots, D\} \setminus \{s\}$}{
        is\_cat = True/False \;
        X\_list $\gets [X]$\;
        J\_list $\gets [J]$\;
        \For{$t = 1$ to $T$}{
            L, position, X\_split, J\_split $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
            splits[s, c, t] $\gets$ L\;
            positions[s, c, t] $\gets$ position\;
        }
    }
}
\end{algorithm}

\begin{algorithm}
\caption{BestSplit}
\label{alg:single_feature_subregion}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

\Input{X\_list, J\_list, s, c, is\_cat}
\Output{BestSplits}

% Initialize variables
\BlankLine
positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
L $\gets$ []\; % \Comment{List to store interaction values}
\For{p \textnormal{in} positions}{
    X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
    L $\gets$ GetInteraction(X\_split, J\_split)\;
}
split\_position = argmin(L)\;
position $\gets$ positions[split\_position]\;
X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, position, is\_cat)\;
\Return{L, position, X\_split, J\_split}
\BlankLine
\end{algorithm}

\begin{algorithm}
\caption{SplitDataset}
\label{alg:split_dataset}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\Input{X\_list, J\_list, c, val, is\_cat}
\Output{X\_split, J\_split}
X\_split $\gets$ []\;
J\_split $\gets$ []\;
\For{i = 1 to len(X\_list)}{
    X= X\_list[i]\;
    J= J\_list[i]\;
    \uIf{is\_cat}{
        ind\_1 $\gets$ X[:, c] $=$ val\;
        ind\_2 $\gets$ X[:, c] $\neq$ val\;
    }
    \Else{
        ind\_1 $\gets$ X[:, c] $\leq$ val\;
        ind\_2 $\gets$ X[:, c] $>$ val\;
    }
    Append X[ind\_1], X[ind\_2] to X\_split\;
    Append J[ind\_1], J[ind\_2] to J\_split\;
}
\Return X\_split, J\_split
\end{algorithm}

\begin{algorithm}
\caption{GetInteraction}
\label{alg:get_interaction}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\Input{$X\_list$, $J\_list$, $min\_points$}
\Output{weighted average of interaction levels}
N $\gets$ total number of items in $X\_list$\;
W $\gets$ []\;
L $\gets$ []\;
\For{$i$ in len($X\_list$)}{
    X $\gets$ X\_list[i]\;
    J $\gets$ J\_list[i]\;
    \uIf{$|X|$ < min\_points}{
      Append $\infty$ to L\;
    }
    \Else{
        Append $\mathcal{L}(X, J)$ to $L$\;
    }
    Append $\frac{|X|}{N}$ to $W$\;
}
\Return $\text{sum}(L \cdot W)$\;
\end{algorithm}

@@

\section{Appendix}

    \subsubsection{Algorithmic Details of Subregion Detection}
